Accessing biological data with semantic web technologies

### 2013-12-30 article.tex
A working remote SPARQL example

Or output is as a Notation3 string with:

@@

...

%\subsection{Example: BioPortal}
%
%In the BRCA1 SNP example, we found ontology terms for concepts using BioPortal. Now, at the
%time of writing BioPortal has an experimental SPARQL end point which allows us to find terms
%in this way too. For example, to find terms related to SNPs, we can run this code in R:
%
%\begin{verbatim}
%findSNPterms = paste(
%  "PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>",
%  "SELECT DISTINCT ?term ?label WHERE {",
%  " ?term rdfs:label ?label .",
%  "FILTER (CONTAINS (LCASE(str(?label)), \"snp\"))",
%  "}"
%)
%sparql.remote(
%  endpoint="http://sparql.bioontology.org/sparql",
%  sparql=findSNPterms, jena=FALSE
%)
%\end{verbatim}

@@

The rrdflibs and rrdf packages bring semantic web technologies to R statistical environment.
While only a subset of Apache Jena is currently exposed, it provides key methods to deal with
RDF data and resources.

>>>>

Or output it as a Notation3 string with:

@@

Remote SPARQLing can be used to get more information about biological entities of interest. The following
example retreives information from the Uniprot database using the experimental SPARQL end point. In
particular, it retrieves annotations for proteins encoded by the BRCA1 gene:

\begin{verbatim}
findAnnotations = paste(
  "prefix up: <http://purl.uniprot.org/core/>",
  "prefix ens: <http://purl.uniprot.org/ensembl/>",
  "prefix rdfs: <http://www.w3.org/2000/01/rdf-schema#>",
  "SELECT DISTINCT ?protein ?mnemonic ?comment WHERE {",
  "  ?transcript up:transcribedFrom ens:ENSG00000012048 .",
  "  ?protein a up:Protein ;",
  "    rdfs:seeAlso ?transcript ;",
  "    up:mnemonic ?mnemonic ;",
  "    up:annotation ?annot .",
  "  ?annot rdfs:comment ?comment .",
  "}"
)
annotations = sparql.remote(
  endpoint="http://beta.sparql.uniprot.org/",
  sparql=findAnnotations, jena=FALSE
)
\end{verbatim}

@@

The rrdflibs and rrdf packages bring basic semantic web technologies to R statistical environment.
While only a subset of Apache Jena is currently exposed, it provides key methods to deal with
RDF data and resources. The examples shed some light on how it can find a place on more advanced
analyses workflows.
